Simple connectome inference from partial correlation statistics in calcium imaging

### 2014-06-19 article.tex
Wordings (1+2)

The human brain is a complex biological organ made of 100 billions of neurons,
each connected to 7000 other neurons on average. Unfortunately, direct
observations of the connectome, the wiring diagram of the brain, is not yet
technically feasible. Without being perfect, calcium imaging currently allows
the realtime and simultaneous observation of neuron activities from thousands
of neurons, producing individual time series representing their fluoresence
intensity. In this context,  the connectome inference problem consists in
retrieving the synaptic connections between neurons on the basis of the
fluoresence time series. In particular, this problem is often made difficult
because of experimental issues, including masking effects making some of the
neurons not to be observed or confounded with others, a low sampling rate of
the optical device with respect to the neural activity speed or the slow decay
of fluoresence.

@@

In this paper, we present a simplified -- yet nearly as good -- version of
the winning method\footnote{Code available at \url{https://github.com/asutera
/kaggle-connectomics}} of the Connectomics
Challenge\footnote{\url{http://connectomics.chalearn.org}}, as a simple and
theoretically grounded approach based on signal processing techniques and
partial correlation statistics. The rest of the paper is structured as
follows: Section \ref{sec:filter} describes the signal processing methods
applied on fluorescent calcium time series; Section \ref{sec:inference} then
presents the proposed approach and its theoretical properties; Section
\ref{sec:results} provides an empirical analysis and comparison with other
network inference methods, while Section \ref{sec:conclusion} finally
discusses our work and provides further research directions. Additionnaly, Appendix
\ref{app:optimized} further describes our actual winning method, giving slightly
better results than the method presented in the paper but highly tuned for the
challenge.

@@

\begin{figure}
\centering
\includegraphics[width=0.4\linewidth]{images/fig_filtering}
\caption{Signal processing process.}
\label{fig:filtered-signal}
\end{figure}

As the first plot in Figure~\ref{fig:filtered-signal} illustrates, due to
light scattering artifacts that ordinarily affects the quality of the
recording~\citep{lichtman2011big}, the raw fluoresence signal is very noisy.
Accordingly, the first step of our pipeline is to smoothen the signal, using a low-pass filter for
filtering out high frequency noise, applying one of the following designs and
resulting in the second plot in the figure:
\begin{align}
% symmetrical median filter
f_1(x^t_i) &= x^{t-1}_i + x^{t}_i + x^{t+1}_i \label{eq:symetric-median}, \\
% asymmetrical weighted median
f_2(x^t_i) &= 0.4 x^{t-3}_i + 0.6 x^{t-2}_i + 0.8 x^{t-1}_i + x_{t}^i.
\label{eq:weighted-asymetric-median}
\end{align}

@@

The full signal processing of our approach is defined by the composed function
$
w \cdot h \cdot g \cdot f_k
$
where $f_k$ is one of the low-pass filter designs.

>>>>

The human brain is a complex biological organ made of 100 billions of neurons,
each connected to 7000 other neurons on average. Unfortunately, direct
observations of the connectome, the wiring diagram of the brain, is not yet
technically feasible. Without being perfect, calcium imaging currently allows
the realtime and simultaneous observation of neuron activities from thousands
of neurons, producing individual time series representing their fluoresence
intensity. From these data, the connectome inference problem consists in
retrieving the synaptic connections between neurons on the basis of the
fluoresence time series. In particular, this problem is often made difficult
because of experimental issues, including masking effects (making some of the
neurons not to be observed or confounded with others), the low sampling rate of
the optical device with respect to the neural activity speed or the slow decay
of fluoresence.

@@

In this paper, we present a simplified -- yet nearly as good -- version of the
winning method\footnote{Code available at \url{https://github.com/asutera
/kaggle-connectomics}} of the Connectomics
Challenge\footnote{\url{http://connectomics.chalearn.org}}, as a simple and
theoretically grounded approach based on signal processing techniques and
partial correlation statistics. The paper is structured as follows: Section
\ref{sec:filter} describes the signal processing methods applied on fluorescent
calcium time series; Section \ref{sec:inference} then presents the proposed
approach and its theoretical properties; Section \ref{sec:results} provides an
empirical analysis and comparison with other network inference methods, while
Section \ref{sec:conclusion} finally discusses our work and provides further
research directions. Additionnaly, Appendix \ref{app:optimized} further
describes, in all details, our actual winning method, giving slightly better
results than the method presented in the paper but highly tuned for the
challenge.

@@

\begin{figure}
\centering
\includegraphics[width=0.4\linewidth]{images/fig_filtering}
\caption{Signal processing process for extracting peaks from the raw fluorescence data.}
\label{fig:filtered-signal}
\end{figure}

As the first plot in Figure~\ref{fig:filtered-signal} illustrates, the raw
fluoresence signal is very noisy due to light scattering artifacts that
ordinarily affects the quality of the recording~\citep{lichtman2011big}.
Accordingly, the first step of our pipeline is to smoothen the signal, using a
low-pass filter for filtering out high frequency noise and resulting in the
second plot in the figure:
\begin{align}
% symmetrical median filter
f_1(x^t_i) &= x^{t-1}_i + x^{t}_i + x^{t+1}_i \label{eq:symetric-median}, \\
% asymmetrical weighted median
f_2(x^t_i) &= 0.4 x^{t-3}_i + 0.6 x^{t-2}_i + 0.8 x^{t-1}_i + x_{t}^i,
\label{eq:weighted-asymetric-median}
\end{align}
where $f_1$ is the filter proposed in the simplified variant of our approach
and $f_2$ is the filter actually used within our winning algorithm.

@@

In those terms, the full signal processing pipeline of our approach is defined
by the composed function $w \cdot h \cdot g \cdot f_1$ (resp $f_2$).
